README.md

### 2024-04-26 README.Rmd
readme

---
title: "Anthropogenic plant communities of the Iberian Atlantic ecoregion"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

Running title: Iberian Atlantic anthropogenic vegetation

Eduardo Fernández-Pascual^1,a^ (<https://orcid.org/0000-0002-4743-9577>), Víctor González-García^1^ (<https://orcid.org/0000-0002-8949-7943>), Greta Ivesdal^2^, Adrián Lázaro-Lobo^1^ (<https://orcid.org/0000-0002-0509-2085>), Borja Jiménez-Alfaro^1^ (<https://orcid.org/0000-0001-6601-9597>)

^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Spain; ^2^ Jardín Botánico Atlántico, Gijón/Xixón, Spain

Correspondence: Eduardo Fernández-Pascual, Instituto Mixto de Investigación en Biodiversidad, Campus de Mieres, Edificio de Investigación, 5ª planta, c/ Gonzalo Gutiérrez Quirós s/n, E-33600 Mieres, Spain. Email: [fernandezpeduardo\@uniovi.es](mailto:fernandezpeduardo@uniovi.es){.email}. Telephone: +34985104781.

# Author contributions

Eduardo Fernández-Pascual: Conceptualization; Methodology; Data Curation; Formal Analysis; Visualization; Writing – Original Draft Preparation; Writing – Review & Editing. Víctor González-García: Methodology; Investigation; Formal Analysis; Writing – Review & Editing. Greta Ivesdal: Investigation; Formal Analysis; Writing – Review & Editing. Adrián Lázaro-Lobo: Investigation; Writing – Review & Editing. Borja Jiménez-Alfaro: Methodology; Writing – Review & Editing.

# Funding

Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico (SV-23-GIJON-JBA).

# Conflict of interest statement

None declared.

# Data availability statement

Original datasets, R code for analysis and Rmarkdown code for creation of the manuscript can be accessed at the GitHub repository <https://github.com/efernandezpascual/manmade>. Upon publication, a version of record of the repository will be deposited in Zenodo.

# Abstract (300 words max.)

@@

* *Results* We classified 2,081 vegetation plots into 28 anthropogenic alliances representing 9 vegetation classes (*Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*). The plots included 1,162 vascular plant taxa: 78 % natives, 15 % archaeophytes and 7 % neophytes. The main ecological gradients behind this diversity went from dry-sunny to moist-shady habitats, followed by a gradient of low to high disturbance frequency and severity.

@@

# Keywords (10 max)

Synanthropic vegetation, human-made habitats, urban biodiversity, neophytes, archaeophytes, alien plants, biological invasions, habitat classification, semi-supervised classification

# Introduction

Anthropogenic vegetation is the set of plant communities that occur as a direct consequence of human activities, and is chiefly composed by the weed vegetation of arable fields and the ruderal vegetation of human settlements and their surroundings [@RN5441]. The anthropogenic communities of human-made habitats make up an important part of the supply of natural elements in urban and peri-urban habitats [@RN5445]. In densely-populated city landscapes, the plant communities that colonize urban vacant lots [@RN2300] are a useful biodiversity resource with high ecological and societal potential [@RN5446]. Anthropogenic vegetation cover can also impact regional-level extreme climatic events [@RN5440]. Furthermore, many synanthropic plants can accumulate heavy metals in their biomass [@RN5433], making them valuable candidates for nature-based solutions such as bioremediation and restoration of degraded post-industrial landscapes [@RN5447].

The anthropogenic vegetation of temperate Europe is formed by an original pool of native species favored by humans, but this pool has been constantly enriched by the arrival of alien species [@RN5504]. First, by archaeophytes, whose arrival started in the Neolithic and peaked in the Bronze Age, with successive increases during periods of accelerated human colonization such as the early Middle Ages [@RN5504]. Synanthropic archaeophytes include many Mediterranean taxa originated within the segetal flora of the Near East [@RN5494]. Some of these archaeophytes are 'obligate weeds' that do not have non-anthropogenic habitats and that are assumed to have evolved together with human agriculture [@RN5494]. Archaeophytes started their expansion into temperate areas during the Mid-Holocene warm period, but it is difficult to disentangle to what extent this expansion was favored by human activities and/or ongoing climatic changes [@RN5432; @RN5443], especially in southern European areas close to the transition between the Mediterranean and the temperate climates [@RN5444]. The synanthropic flora of Europe has been further enriched by neophytes, alien taxa that arrived from the start of the Modern Age (after 1.500 A.D.) to the present day [@RN5439]. Synanthropic neophytes include many American, African and Asian taxa that got to Europe as a consequence of the global trade networks established by the Western European colonial powers [@RN5451]. In the last century, the European anthropogenic vegetation has shown a decrease in species richness and diversity [@RN5423], as rare synanthropic species have become rarer, and neophytes have increased their abundance at the expense of natives and archaeophytes [@RN5441].

Anthropogenic vegetation is likely to be similar within ecological regions (ecoregions, hereafter), since they are assumed to encompass areas with a similar biogeographic history and recurrent local ecosystems [@RN5491]. However, ecoregional-scale studies on the diversity of anthropogenic vegetation are scarce in the literature. The Iberian Atlantic ecoregion (a.k.a. Cantabrian Mixed Forests ecoregion; NW Iberian Peninsula) has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. This, together with its transitional position at the border between the temperate and Mediterranean climatic zones of Europe [@RN4700], suggests a long regional history for archaeophyte-rich anthropogenic plant communities, likely supporting a current high diversity of anthropogenic plant communities with both temperate and Mediterranean floristic elements [@RN5008]. Historical trade links with America and Asia, together with the warm and humid temperate climate, have made the region a hotspot for neophyte invasions [@RN5492; @RN5493]. Recent post-industrial land-use changes have left large areas with abandoned industrial sites (i.e. brownfields) whose management and restoration requires an understanding of anthropogenic communities able to colonize or restore them [@RN5448; @RN5449]. Although there is a long tradition of studying weed and ruderal communities in the ecoregion [@RN5364; @RN5361; @RN5359; @RN5360] there is a current need for an ecoregional-level synthesis that revises and updates the classification in accord with recent developments in European vegetation science [@RN4698] and habitat classification [@RN5061]. In this article, we have performed such a synthesis with the following objectives: (1) provide an updated classification of anthropogenic plant communities in the Iberian Atlantic ecoregion; (2) quantify the proportion of natives, archaeophytes and neophytes in different anthropogenic plant communities; (3) characterize the plant communities in terms of life forms, plant height and flowering phenology; (4) describe the ecological requirements of the different communities (temperature, moisture, light, nutrients, soil reaction, disturbance frequency and disturbance severity); and (5) provide a summary description of each of the plant communities.

# Methods

## Data curation and taxonomic criteria

We performed all data management and analysis with R version 4.3.1 [@RN5387], using the R package *tidyverse* [@RN4662] for data processing and visualization. We homogenized all taxon names in the analysis and the article using Euro+Med [@RN5044] or Plants of the World Online [@RN5376] for taxa not in Euro+Med. For the nomenclature of sintaxa, we followed Mucina *et al.* [-@RN4698] for alliances and higher ranks, and Rivas-Martínez *et al.* [-@RN4697] for associations. The original datasets, as well as R code for analysis and creation of the manuscript, can be accessed at the GitHub repository <https://github.com/efernandezpascual/manmade>.

## Study ecoregion

We studied the anthropogenic plant communities of the Iberian Atlantic ecoregion [@RN5372], i.e. the territories with a temperate climate in the north-western part of the Iberian Peninsula. Our study ecoregion broadly corresponds with the Cantabrian Mixed Forests ecoregion *sensu* Olson *et al.* [-@RN4667], with the Iberian part of the European Atlantic province *sensu* Rivas-Martínez *et al.* [-@RN4699] and with the Iberian section of the Atlantic biogeographical region of the European Environmental Agency (<https://www.eea.europa.eu/data-and-maps/figures/biogeographical-regions-in-europe-2>).

## Definition of anthropogenic vegetation

To circumscribe our study vegetation, we followed the definition of anthropogenic vegetation in the recent classification of the vegetation of Europe by Mucina *et al.* [-@RN4698]. In our study area, this potentially includes the vegetation classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*. To these, and for a complete synthesis, we added the class *Cymbalario-Parietarietea diffusae* included by Mucina *et al.* [-@RN4698] in the vegetation of rock crevices and screes, since this class encompasses the vegetation of human-made walls.

## Checklist of anthropogenic sintaxa

To assist in our classification, we prepared a checklist of anthropogenic sintaxa that could be present in our study ecoregion (**Appendix S1**), based on regional sintaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. The sintaxonomical checklist included 38 anthropogenic vegetation alliances that could be present in the Iberian Atlantic ecoregion. A goal of our study was to validate the presence of these alliances in the ecoregion, and to assign to them the existing vegetation data.

## Vegetation data selection

We retrieved Iberian Atlantic anthropogenic vegetation plots from SIVIM, the Iberian and Macaronesian Vegetation Information System [@RN5374]. To identify plots as anthropogenic, we selected those plots that had been classified by the original authors of the plot as belonging to any sintaxa in our sintaxonomical checklist. However, since not all plots in SIVIM have an assigned sintaxon, we also retrieved additional plots using the expert system created by Chytrý *et al.* [-@RN5061] to classify vegetation plots into EUNIS pan-European habitat types. We retrieved plots that had been assigned to any habitat related to the vegetation classes we had previously defined: all habitats in the level 1 code *V Vegetated man-made habitats*, plus level 3 codes *R55 Lowland moist or wet tall-herb and fern fringe* and *R57 Herbaceous forest clearing vegetation*. It must be noted that habitats *R55* and *R57* include communities that are classified as *R Grasslands and lands dominated by forbs, mosses or lichens* by EUNIS but as *Anthropogenic vegetation* by Mucina *et al.* [-@RN4698]. The extraction from SIVIM produced an initial pool of 3,160 vegetation plots, to which we added 89 vegetation plots of urban or peri-urban plant communities sampled by us or extracted from local literature not in SIVIM [@RN5377; @RN5378].

## Vegetation data cleaning

We performed an exploratory analysis of these 3,249 plots using modified Two-Way Indicator Species Analysis (TWINSPAN) [@RN5379]. We used the R package *twinspanR* [@RN5380] for TWINSPAN, with 3 pseudospecies cut levels (0, 15, 25), a minimum group size of 10 plots and Sørensen's average dissimilarity. During this exploratory analysis we identified 741 outlier plots which did not belong to either the target anthropogenic vegetation or to the geographical area of the ecoregion. The majority of these plots corresponded to coastal plant communities and had been misclassified by the expert system. We removed these, leaving a dataset of 2,508 vegetation plots for data analysis: 2,419 plots from SIVIM, originally recorded in 89 publications (**Appendix S2**); plus 89 plots added by us.

## Training dataset for semi-supervised classification and validation of alliances

To create a training dataset for semi-supervised classification (see below), we started by keeping only those plots (n = 2,201) that had been assigned by the original authors of the plot as belonging to any of the 38 alliances in our sintaxonomical checklist (i.e., we removed the plots without sintaxa, that had been retrieved solely by the expert system). We further cleaned this subset by subjecting it to a preliminary numerical classification using TWINSPAN, in which we attempted to match the plots into the 38 alliances in our sintaxonomical checklist. This step allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification and (2) the classification based on the original author's sintaxon; these plots (n = 1,725) became our training dataset for semi-supervised classification. Furthermore, this step allowed us to validate which alliances from our sintaxonomical checklist were present in the study ecoregion (n = 28) and which were absent (n = 10).

## Semi-supervised classification

Next, we conducted a semi-supervised classification of the whole dataset (n = 2,508 plots) into the 28 validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance, and thus we kept the 28 alliances as final vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 plots into 28 alliances and 427 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the classes and alliances.

## Characteristic species and EUNIS habitat correspondence

To facilitate the usage of the classification by local managers and environmental consultants, we calculated sets of characteristic species (**Appendix S3**) [@RN5061] for each of the final 28 vegetation alliances. We defined dominant species as species with more than 25% cover in at least 5% of the vegetation plots of the group; constant species as species with a frequency higher than 50% in the group; and diagnostic species as species with an *IndVal* value with a p-value lower than 0.05, as calculated with the R package *labdsv* [@RN2311] using 1.000.000 iterations. Additionally, we assigned to each alliance its correspondence to level-3 EUNIS habitat codes (<https://eunis.eea.europa.eu/habitats-code-browser-revised.jsp>).

## Species origin as native, archaeophyte and neophyte

We classified the species as native, archaeophytes and neophytes in the Iberian Atlantic ecoregion using the information in *Flora iberica* [-@RN3243] and catalogues of archaeophytes for Britain [@RN5490] and the Czech Republic [@RN5495]. We must stress that identifying archaeophytes in southern Europe is highly problematic and our classification must be taken as an indication of putative archaeophyte character for the purposes of vegetation description, rather than a definitive classification of the species: as stated by Preston *et al.* [@RN5490], the classification of a species as an archaeophyte should be interpreted as a hypothesis to be tested by further studies.

## Species life form

From *FloraVeg.EU* [-@RN5385] we extracted the species’ life forms and kept *therophyte* and *geophyte* for further analysis since these two categories had the largest contribution to variance as per an exploratory Principal Component Analysis (PCA) performed using the R package *FactoMineR* [@RN3166].

## Species height and flowering phenology

From *Flora iberica* [-@RN3243] we extracted the maximum height, median month of flowering and length of the flowering period of each species in the dataset. We used these variables to calculate plot-level community-weighted means (weighting by species cover) for each trait, to characterize the vegetation height and flowering phenology of the anthropogenic plant communities.

## Species ecological indicator values

We collected, for each species in the dataset, its ecological indicator values of temperature, moisture, light, nutrients and soil reaction [@RN5384]; and its disturbance frequency and severity indicator values [@RN5101]. To those species lacking an indicator value in the accessed references, we assigned a value by reciprocal averaging, i.e. by (1) calculating weighted average values of the species with values for each plot (weighting by species cover), and (2) assigning to the missing species the weighted average values of the plots where they were present (weighting by species cover). Then, for each vegetation plot, we calculated the plot-level mean of each indicator value.

# Results

## Overview of the classification

The semi-supervised classification resulted in the classification of 2,081 vegetation plots into 28 anthropogenic alliances (**Fig. 1**) representing 9 vegetation classes. Most of the alliances had a wide distribution in the ecoregion according to our dataset, but five of them (*Caucalidion lappulae*, *Linario polygalifoliae-Vulpion alopecuri*, *Convolvulo arvensis-Agropyrion repentis*, *Senecionion fluviatilis* and *Paspalo-Agrostion semiverticillati*) had isolated occurrences in a few points towards the southern limit of the Iberian Atlantic territories. The regional literature included in our dataset recognized 72 associations (**Appendix S4**) within these alliances, but exploratory analysis (not shown) indicated that the separation of these associations within the alliances had a generally weak support based on numerical classification methods.

The classified plots included 1,162 taxa or taxa aggregates. The 10 most frequent species were *Ochlopoa annua* (621 occurrences), *Urtica dioica* (592), *Sonchus oleraceus* (540), *Stellaria media* (516), *Capsella bursa-pastoris* (390), *Polygonum aviculare* (375), *Dactylis glomerata* (363), *Senecio vulgaris* (332) and *Anisantha sterilis* (301). Considering only those plots with the most frequent plot size (20 m^2^, n = 256 plots), the average species richness per plot was 16 (minimum = 3, maximum = 37). The class with the richest species pool was *Epilobietea angustifolii* (n = 615) and the poorest was *Cymbalario-Parietarietea diffusae* (n = 171).

PCA ordination of the floristic composition of the classified plots (**Fig. 2A**) suggested a high consistency of three major vegetation groups: (1) trampled vegetation of class *Polygono-Poetea annuae*; (2) crop weeds of classes *Papaveretea rhoeadis* and *Digitario sanguinalis-Eragrostietea minoris*; and (3) perennial ruderal vegetation of classes *Artemisietea vulgaris* and *Epilobietea angustifolii*. The central position of the floristic space, between these three major groups, was occupied by the annual ruderal vegetation of classes *Chenopodietea*, *Sisymbrietea* and *Bidentetea*; as well as by the wall vegetation of class *Cymbalario-Parietarietea diffusae*. Separate PCAs within the major vegetation groups indicated a coherent separation between most of the alliances, but with some cases of relatively high overlap: for example, between (i) *Galio valantiae-Parietarion judaicae* and *Cymbalario-Asplenion*; (ii) *Geo urbani-Alliarion officinalis* and *Aegopodion podagrariae*.

## Proportion of natives, archaeophytes and neophytes

The total synanthropic species pool (n = 1,162) was composed of 78 % natives (n = 908), 15 % putative archaeophytes (n = 178) and 7 % neophytes (n = 76). The most frequent putative archaeophytes were *Capsella bursa-pastoris* (390 occurrences), *Senecio vulgaris* (332), *Anisantha sterilis* (301), *Anthemis arvensis* (280), *Hordeum murinum* (274), *Raphanus raphanistrum* (233), *Vicia sativa* (228), *Malva sylvestris* (197), *Sisymbrium officinale* (166) and *Echinochloa crus-galli* (165); while the most frequent neophytes were *Veronica persica* (212), *Erigeron canadensis* (138), *Amaranthus hybridus* (73), *Oxalis corniculata* (67), *Lepidium didymum* (57), *Oxalis latifolia* (51), *Paspalum distichum* (48), *Bidens frondosus* (35), *Cyperus eragrostis* (34) and *Symphyotrichum squamatum* (33).

By vegetation class, *Epilobietea angustifolii* had the highest proportion of natives in its species pool (82 %); *Sisymbrietea* had the highest proportion of putative archaeophytes (31 %); and *Bidentetea* had the highest proportion of neophytes (17 %). The alliances with the highest proportions of archaeophytes in their species pools were *Caucalidion lappulae* (41 %), *Chenopodion muralis* (35 %) and *Sisymbrion officinalis* (31 %); while the alliances with the highest proportion of neophytes were *Bidention tripartitae* (19 %), *Allion triquetri* (12 %) and *Spergulo arvensis-Erodion cicutariae* (10%).

Focusing on plot-level average proportions (**Fig. 3**), the alliances with the most archaeophytes were *Caucalidion lappulae*, *Spergulo arvensis-Erodion cicutariae* and *Sisymbrion officinalis*, in which only about half of the plot-level proportion was made up of native species. The alliances with the highest plot-level proportion of neophytes were *Bidention tripartitae*, *Paspalo-Agrostion semiverticillati* and *Senecionion fluviatilis*. Some alliances had high proportions of native species, e.g. *Epilobion angustifolii*, *Convolvulo arvensis-Agropyrion repentis* and *Linario polygalifoliae-Vulpion alopecuri*.

## Community-level traits

The most frequent life form in the species pool (n = 1,162 species) were hemicryptophytes (44 %), followed by therophytes (42 %), geophytes (9 %), chamaephytes (6 %), phanerophytes (6 %), bryophytes (1 %) and hydrophytes (1 %). The proportion of therophytes across classes and alliances (**Fig. 4**) agreed with the traditional description of the sintaxa. Thus, therophytes dominated the annual communities of trampled-soil vegetation (*Polygono-Poetea annuae*), crops weeds (*Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*) and annual ruderals (*Chenopodietea*, *Sisymbrietea*, *Bidentetea*); while perennial life forms (especially hemicryptophytes) dominated walls (*Cymbalario-Parietarietea diffusae*) and perennial ruderal vegetation (*Artemisietea vulgaris*, *Epilobietea angustifolii*). Geophytes were a relatively high proportion of the vegetation in three alliances: *Allion triquetri*, *Convolvulo arvensis-Agropyrion repentis* and *Senecionion fluviatilis*.

Community-weighted means for plant height (**Fig. 5A**) also agreed with the expected description of the sintaxa. The shortest communities were the dwarf-herb vegetation of trampled sites (*Polygono-Poetea annuae*, average height = 45.2 cm) and crop weeds (*Papaveretea rhoeadis*, 55.4 cm), while the tallest were the perennial ruderal vegetation of dry (*Artemisietea vulgaris*, 129 cm) and mesic (*Epilobietea angustifolii*, 136 cm) sites. The median month of flowering (**Fig. 5B**) was May (*Sisymbrietea*, *Chenopodietea*, *Papaveretea rhoeadis*), June (*Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Epilobietea angustifolii*, *Artemisietea vulgaris*) or July (*Digitario sanguinalis-Eragrostietea minoris*, *Bidentetea*). The length of the flowering season (**Fig. 5C**) was generally high, from 5 months (*Artemisietea vulgaris*, *Epilobietea angustifolii*) to 6 (*Chenopodietea*, *Sisymbrietea*), 7 (*Bidentetea*, *Digitario sanguinalis-Eragrostietea minoris*, *Papaveretea rhoeadis*), 8 (*Polygono-Poetea annuae*) or 9 (*Cymbalario-Parietarietea diffusae*).

## Community-level ecological preferences

To visualize the major patterns of variation in the ecological and disturbance preferences of the different communities (**Fig. 6**) we performed a PCA ordination of plot-level means (**Fig. 7**). The first two PCA axes explained 59 % of the variability. The major contributors to axis 1 (36 % variance explained) were light, nutrient and moisture requirements; it separated communities of open, dry and comparatively nutrient-poorer sites (classes *Papaveretea rhoeadis*, *Polygono-Poetea annuae*, *Sisymbrietea*, *Artemisietea vulgaris*, *Chenopodietea* and *Digitario sanguinalis-Eragrostietea minoris*) from communities of shady, moist and nutrient-richer sites (classes *Cymbalario-Parietarietea diffusae*, *Bidentetea* and *Epilobietea angustifolii*). The major contributors to axis 2 (23% variance explained) were disturbance severity and frequency; this axis separated communities preferring low severity - low frequency disturbances (classes *Cymbalario-Parietarietea diffusae*, *Epilobietea angustifolii*, *Chenopodietea*, *Artemisietea vulgaris* and *Bidentetea*) from communities adapted to high severity - high frequency disturbances (classes *Papaveretea rhoeadis*, *Polygono-Poetea annuae*, *Sisymbrietea* and *Digitario sanguinalis-Eragrostietea minoris*).

## Description of the Iberian Atlantic anthropogenic communities

***Cymbalario-Parietarietea diffusae*** Vegetation of human-made walls.

1.  *Galio valantiae-Parietarion judaicae* Hemicryptophyte-rich vegetation of walls. Occupies slightly warmer and sunnier situations than the other alliance in the class and has a shorter flowering season. Wide distribution in the ecoregion. Diagnostic species in the ecoregion: *Parietaria judaica*, *Centranthus ruber*, *Asplenium trichomanes*. Correspondence with EUNIS (2012) habitat *J2.5 - Constructed boundaries*.

2.  *Cymbalario-Asplenion* Fern-rich vegetation of walls. Occupies slightly colder and shadier walls than the other alliance in the class and has a longer flowering season. Wide distribution in the ecoregion. Diagnostic species in the ecoregion: *Cymbalaria muralis*. Correspondence with EUNIS (2012) habitat *J2.5 - Constructed boundaries*.

***Polygono-Poetea annuae*** Dwarf-herb vegetation of heavily-trampled sites.

3.  *Polycarpion tetraphylli* Dwarf-annual trampled vegetation of warm and sunny sites. Occupies warmer, sunnier and nutrient-poorer situations than the other alliances in the class and has an earlier flowering season. Wide distribution in the ecoregion, specially along the coast and in submediterranean valleys of the interior. Diagnostic species in the ecoregion: *Sagina apetala*, *Plantago coronopus*, *Crassula tillaea*, *Spergularia marina*. Correspondence with EUNIS habitat *V34 Trampled xeric grassland with annuals*.

4.  *Polygono-Coronopodion* Annual trampled vegetation of dry sites. Occupies sites that are colder than those favored by *Polycarpion tetraphylli*, but drier than those of *Saginion procumbentis*. Has a specially high proportion of neophytes. Wide distribution in the ecoregion. Diagnostic species in the ecoregion: *Matricaria suaveolens*, *Polygonum aviculare*. Correspondence with EUNIS habitat *V34 Trampled xeric grassland with annuals*.

5.  *Saginion procumbentis* Bryophyte-rich trampled vegetation of mesic sites. Occupies colder and wetter situations than the other alliances in the class, is less dominated by therophytes and has a relatively high proportion of bryophytes. Distribution concentrated in the central part of the ecoregion. Diagnostic species in the ecoregion: *Sagina procumbens*, *Bryum argenteum*. Correspondence with EUNIS habitat *V35 Trampled mesophilous grassland with annuals*.

***Papaveretea rhoeadis*** Annual weed vegetation of crops and gardens.

6.  *Caucalidion lappulae* Annual weed vegetation of cereal crops on base-rich soils. Very high proportion of putative archaeophytes. Distribution mostly limited to the more submediterranean valleys south of the Cantabrian Mountains. Diagnostic species in the ecoregion: *Cyanus segetum*, *Papaver rhoeas*, *Trifolium arvense*, *Anacyclus clavatus*, *Valerianella eriocarpa*. Correspondence with EUNIS habitats *V11 Intensive unmixed crops* and *V13 Arable land with unmixed crops grown by low-intensity agricultural methods*.

7.  *Scleranthion annui* Annual weed vegetation of cereal crops on base-poor soils. Occupies more acidic soils than the other alliances in the class, and flowers earlier. High proportion of putative archaeophytes. Wide distribution in the ecoregion, but specially in the west, where acidic bedrocks dominate. Diagnostic species in the ecoregion: *Mibora minima*, *Rumex acetosella*. Correspondence with EUNIS habitats *V11 Intensive unmixed crops*, *V12 Mixed crops of market gardens and horticulture*, *V13 Arable land with unmixed crops grown by low-intensity agricultural methods* and *V15 Bare tilled, fallow or recently abandoned arable land*.

8.  *Oxalidion europeae* Annual weed vegetation of gardens and root crops. Occupies wetter, shadier and nutrient-richer situations than the other alliances in the class, has a longer flowering season. High proportion of putative archaeophytes, and more neophytes than the other alliances in the class. Wide distribution in the ecoregion. Diagnostic species in the ecoregion: *Stellaria media*, *Veronica persica*. Correspondence with EUNIS habitats *V11 Intensive unmixed crops*, *V12 Mixed crops of market gardens and horticulture*, *V13 Arable land with unmixed crops grown by low-intensity agricultural methods* and *V15 Bare tilled, fallow or recently abandoned arable land*.

***Digitario sanguinalis-Eragrostietea minoris*** Summer-annual C4 weed vegetation.

9.  *Spergulo arvensis-Erodion cicutariae* Summer-annual C4 weed vegetation. Late-flowering alliance rich in neophytes and grasses, occurring in highly-disturbed and nutrient-rich sites. Wide distribution in the ecoregion. Diagnostic species in the ecoregion: *Digitaria sanguinalis*, *Echinochloa crus-galli*, *Chenopodium album*, *Amaranthus hybridus*. Correspondence with EUNIS habitats *V11 Intensive unmixed crops*, *V12 Mixed crops of market gardens and horticulture* and *V13 Arable land with unmixed crops grown by low-intensity agricultural methods*.

***Chenopodietea*** Winter-annual fringe and ruderal vegetation.

10. *Allion triquetri* Geophyte-rich fringe vegetation. This vegetation occupies warmer, shadier and less frequently disturbed situations than other alliances in the class, and is comparatively taller and earlier-flowering. Has a higher proportion of geophytes and a lower proportion of therophytes. The proportion of neophytes is the highest in the class. Wide distribution in the ecoregion. Diagnostic species in the ecoregion: *Urtica membranacea*, *Smyrnium olusatrum*, *Tradescantia fluminensis*. Correspondence with EUNIS habitat *V37 Annual anthropogenic herbaceous vegetation*.

11. *Geranio pusilli-Anthriscion caucalidis* Winter-annual fringe vegetation. Occupies situations similar to the previous alliance, but in colder sites, and is a shorter vegetation dominated by therophytes instead of geophytes. Mostly distributed in the submediterranean valleys south of the Cantabrian Mountains. Diagnostic species in the ecoregion: *Geranium lucidum*, *Anthriscus caucalis*. Correspondence with EUNIS habitat *V37 Annual anthropogenic herbaceous vegetation*.

12. *Chenopodion muralis* Low-growth winter-annual ruderal vegetation. Short vegetation occupying colder situations than the other alliances in the class and more disturbed sites. Flowers longer and later. Has a high proportion of archaeophytes. Sparse occurrence throughout the ecoregion. Diagnostic species in the ecoregion: *Malva neglecta*, *Urtica urens*. Correspondence with EUNIS habitat *V37 Annual anthropogenic herbaceous vegetation*.

13. *Echio-Galactition tomentosae* Tall-herb winter-annual ruderal vegetation. Occupies drier, sunnier and less nutrient-rich situations than other alliances in the class. Mostly distributed in the west of the ecoregion. Diagnostic species in the ecoregion: *Galactites tomentosus*, *Coleostephus myconis*, *Anisantha rigida*. Correspondence with EUNIS habitat *V37 Annual anthropogenic herbaceous vegetation*.

14. *Linario polygalifoliae-Vulpion alopecuri* Ephemeral annual vegetation of disturbed coastal dunes. Occupies warm, dry and sunny coastal sites, and has a high number of coastal dune species. Isolated occurrence in the southern coast of the ecoregion. Diagnostic species in the ecoregion: *Vulpia alopecuros*, *Malcolmia littorea*, *Crucianella maritima*, *Artemisia campestris*. Correspondence with EUNIS habitat *V32 Mediterranean subnitrophilous annual grassland*.

***Sisymbrietea*** Summer-annual ruderal vegetation.

15. *Sisymbrion officinalis* Summer-annual ruderal vegetation. Rich in therophytes and archaeophytes and adapted to severe disturbances. Wide distribution in the ecoregion. Diagnostic species in the ecoregion: *Hordeum murinum*, *Sisymbrium officinale*. Correspondence with EUNIS habitats *V15 Bare tilled, fallow or recently abandoned arable land* and *V37 Annual anthropogenic herbaceous vegetation*.

***Bidentetea*** Summer-annual pioneer vegetation of temporarily flooded sites.

16. *Bidention tripartitae* Summer-annual pioneer vegetation of temporarily flooded sites. Occupies nutrient-richer sites than the other alliance in the class and has a higher proportion of therophytes and neophytes. Dispersed throughout the ecoregion. Diagnostic species in the ecoregion: *Persicaria hydropiper*, *Bidens frondosus*, *Lythrum salicaria*. Correspondence with EUNIS habitat *Q61 Periodically exposed shore with stable, eutrophic sediments with pioneer or ephemeral vegetation*.

17. *Paspalo-Agrostion semiverticillati* Summer-annual pioneer vegetation of temporarily flooded and warm sites. Occupies wetter and more frequently disturbed sites than the other alliance in the class and has a longer flowering season. Isolated occurrences in the south of the ecoregion. Diagnostic species in the ecoregion: *Paspalum distichum*, *Cyperus longus*, *Schoenoplectus lacustris*. Correspondence with EUNIS habitat *R55 Lowland moist or wet tall-herb and fern fringe*.

***Artemisietea vulgaris*** Perennial ruderal vegetation of dry sites.

>>>>

---
title: "README.md"
output: github_document
---

@@

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6077618.svg)](https://doi.org/10.5281/zenodo.6077618)

![Antigua fábrica de La Cerámica Gijonesa](ceramica.JPG)

# Anthropogenic plant communities of the Iberian Atlantic ecoregion

This repository stores all the information related to the manuscript [*Anthropogenic plant communities of the Iberian Atlantic ecoregion*](https://github.com/efernandezpascual/manmade/blob/master/doc/manuscript.md), including the raw datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the manuscript.

## Contents

This repository is organised following the advice of [Wilson et al. 2017](https://doi.org/10.1371/journal.pcbi.1005510) for recording and storing research projects.

The following materials are available in the folders of this repository:

* `data` Data files including the vegetation plots.
* `doc` Files to create the [manuscript](https://github.com/efernandezpascual/manmade/blob/master/doc/manuscript.md) using `Rmarkdown`.
* `results` Output of the `R` scripts, including figures and model outputs.
* `src` Scripts in `R` language used to clean the raw data and perform the analyses of the manuscript.

## Abstract

@@

* *Results* We classified 2,081 vegetation plots into 28 anthropogenic alliances representing 9 vegetation classes (*Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*). The plots included 1,162 plant taxa: 78% natives, 15% archaeophytes and 7% neophytes. The main ecological gradients behind this diversity went from dry-sunny to moist-shady habitats, followed by a gradient of low to high disturbance frequency and severity.

@@

## Citation

Please cite the repository, datasets and article as:

> Fernández-Pascual E, González-García, Ivesdal G, Lázaro-Lobo A, Jiménez-Alfaro B. [*Anthropogenic plant communities of the Iberian Atlantic ecoregion*](https://github.com/efernandezpascual/manmade/blob/master/doc/manuscript.md), *in preparation*.
